Experimentally informed site-specific substitution models deepen phylogenetic estimates of the divergence of viral lineages

### 2018-02-07 manuscript/manuscript.tex
changed model feature figure and add citations

\begin{figure}[H]
\centerline{\includegraphics[width=0.90\textwidth]{figures/model_feature.pdf}}
\caption{\label{fig:model_feature}
\textbf{Different ways of modeling site-to-site variation in purifying selection.}
\jdbcomment{Panels A and B need to be switched to match order in text. Then text updated too}
(A) Substitution model stationary states can either be identical at every site in the protein or allow to vary from site to site. 
(B) The relative rate of non-synonymous change, $\omega$, can be defined as one, gene-wide average or allowed to vary following some statistical distribution.
In order to achieve computational tractability, the distribution is discretized into $K$ bins and $\omega$ is set to the mean of each bin. 
A $\Gamma$ distribution with $K=4$ bins is shown here. 
(C) Substitution models can both, one, or neither of these features. 
We used models from the GY94 and the ExpCM families. 
}
\end{figure}

@@

The most common strategy to model rate variation is to allow the rate of non-synonymous change to vary among sites following some statistical distribution \citep[\ref{fig:model_feature}B;][]{yang1994maximum, yang2000codon}. 
Under such models, some sites have a higher rate of non-synonymous change than others. 
However, this rate is agnostic to the amino-acid identities themselves, and so all non-synonymous changes at at given site are treated equally.

In contrast, so-called ``mutation-selection'' models~\citep{halpern1998evolutionary} account for purifying selection by explicitly defining a different set of amino-acid preferences at each site in the protein. 
This more mechanistic formulation, without an arbitrary statistical distribution, results in a site-specific stationary state (\ref{fig:model_feature}A). 
These models better capture the site-to-site differences in amino-acid composition that is an obvious features of real proteins, and indeed they generally better describe actual evolution than models with only rate variation~\jdbcomment{cite Rodrigue / Lartillot papers, our papers, etc}.

The specificity of mutation-selection models comes at a cost in the form of an increased number of parameters. 
While codon substitution models with uniform stationary states typically have $<$20 parameters, mutation-selection models must specify 19 parameters for \emph{each} site (the stationary state is for 20 amino acids whose frequencies are constrained to sum to one).
This corresponds to $19\times L$ parameters for a protein of length $L$, which leads to $\sim 10^4$ parameters for a typical size protein.
As with all parameter-rich models, it is important to obtain values for these parameters using some method that avoids overfitting.
Here we will primarily use experimentally informed codon models (ExpCM) which define values for these parameters \textit{a priori} from deep mutational scanning experiments so they do not need to be fit from phylogenetic data \jdbcomment{include citations}.
Therefore, for an ExpCM the number of remaining free parameters that are fit from the phylogenetic data are similar to a non-site-specific substitution model.
Alternative strategies of obtaining parameters for site-specific stationary states via Bayesian or maximum-likelihood estimation are discussed in the last section of the Results.

@@

\begin{figure}[H]
\centerline{\includegraphics[width=0.90\textwidth]{figures/decay.pdf}}
\caption{\label{fig:decay}
\textbf{Effect of stationary state and rate variation on long branch estimation.}
The expected pairwise amino-acid identity for five sites in influenza hemagglutinin (HA) for four different substitution models. 
The logoplots show the HA amino-acid preferences from a deep mutational scan performed by \cite{doud2016accurate}. 
The implied dN/dS value was calculated from the ExpCM following \cite{spielman2015relationship}.
}
\end{figure}

@@

\begin{figure}[H]
\centerline{\includegraphics[width=0.9\textwidth]{figures/empirical_trees.pdf}}
\caption{\label{fig:empirical_trees}
\textbf{Trees optimized with an ExpCM defined by H1 preferences lengthen branches from the focal H1 sequence compared to GY94 models.} ).
}
\end{figure}

>>>>

\begin{figure}[H]
\centerline{\includegraphics[width=0.90\textwidth]{figures/model_feature.pdf}}
\caption{\label{fig:model_feature}
\textbf{Different ways of modeling site-to-site variation in purifying selection.}
(A) Substitution model stationary states can either be identical at every site in the protein or allow to vary from site to site. 
(B) The relative rate of non-synonymous change, $\omega$, can be defined as one, gene-wide average or allowed to vary following some statistical distribution.
In order to achieve computational tractability, the distribution is discretized into $K$ bins and $\omega$ is set to the mean of each bin. 
A $\Gamma$ distribution with $K=4$ bins is shown here. 
(C) Substitution models can both, one, or neither of these features. 
We used models from the GY94 and the ExpCM families. 
}
\end{figure}

@@

The most common strategy to model rate variation is to allow the rate of non-synonymous change to vary among sites following some statistical distribution \citep[\ref{fig:model_feature}A;][]{yang1994maximum, yang2000codon}. 
Under such models, some sites have a higher rate of non-synonymous change than others. 
However, this rate is agnostic to the amino-acid identities themselves, and so all non-synonymous changes at at given site are treated equally.

In contrast, so-called ``mutation-selection'' models~\citep{halpern1998evolutionary} account for purifying selection by explicitly defining a different set of amino-acid preferences at each site in the protein. 
This more mechanistic formulation, without an arbitrary statistical distribution, results in a site-specific stationary state (\ref{fig:model_feature}B). 
These models better capture the site-to-site differences in amino-acid composition that is an obvious features of real proteins, and indeed they generally better describe actual evolution than models with only rate variation~\citep{lartillot2004bayesian, lartillot2007suppression, le2008phylogenetic, rodrigue2010mutation,si2008empirical,rodrigue2014site,bloom2017identification,hilton2017phydms,bloom2014experimentally}.
The specificity of mutation-selection models comes at a cost in the form of an increased number of parameters. 
While codon substitution models with uniform stationary states typically have $<$20 parameters, mutation-selection models must specify 19 parameters for \emph{each} site (the stationary state is for 20 amino acids whose frequencies are constrained to sum to one).
This corresponds to $19\times L$ parameters for a protein of length $L$, which leads to $\sim 10^4$ parameters for a typical size protein.
As with all parameter-rich models, it is important to obtain values for these parameters using some method that avoids overfitting.
Here we will primarily use experimentally informed codon models (ExpCM) which define values for these parameters \textit{a priori} from deep mutational scanning experiments~\citep{araya2011,fowler2010high} so they do not need to be fit from phylogenetic data~\citep{bloom2014experimentally, hilton2017phydms, bloom2017identification}.
Therefore, for an ExpCM the number of remaining free parameters that are fit from the phylogenetic data are similar to a non-site-specific substitution model.
Alternative strategies of obtaining parameters for site-specific stationary states via Bayesian\skhcomment{cite} or maximum-likelihood estimation\skhcomment{cite} are discussed in the last section of the Results.

@@

\begin{figure}[H]
\centerline{\includegraphics[width=0.90\textwidth]{figures/decay.pdf}}
\caption{\label{fig:decay}
\textbf{Effect of stationary state and rate variation on long branch estimation.}
The expected pairwise amino-acid identity for five sites in influenza hemagglutinin (HA) for four different substitution models. 
The logoplots show the HA amino-acid preferences from a deep mutational scan performed by \cite{doud2016accurate}. 
The implied dN/dS value was calculated from the ExpCM following \cite{spielman2015relationship}.
\skhcomment{calculated using equation. Maybe change long branch length to expected asymptotic sequence divergence.}
}
\end{figure}

Given a single branch, a substitution model transforms sequence divergence into branch length. 
Under the molecular clock assumption, this branch length is proportional to time. 
However, every substitution model has a stationary state, defining the expected sequence composition after a long evolutionary time. 
This stationary state can be used to calculate the expected amino-acid divergence a very long amount of time has elapsed. 
(Rate only affects how fast asymptotic divergence approached, whereas). 
In \ref{fig:decay}, the stationary state is represented by the ``long" tails where the expected sequence divergence remains constant as time increases.

This is a fundamental feature. 
The stationary state is principle eigenvector of substitution matrix. 
Rate variation only multiples by a number, and stationary state of a stochastic matrix is invariant to multiplication.

@@

\begin{figure}[H]
\centerline{\includegraphics[width=0.8\textwidth]{figures/empirical_trees.pdf}}
\caption{\label{fig:empirical_trees}
\textbf{Trees optimized with an ExpCM defined by H1 preferences lengthen branches from the focal H1 sequence compared to GY94 models.} ).
}
\end{figure}
